---
title: "Analytics Programming Workflow"
author: "Nicholas Jojola"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    theme: cosmo
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(ggplot2)
library(DiagrammeR)
library(knitr)
library(kableExtra)
```
## Project Overview

##### This project demonstrates an analytics workflow implemented in R for cleaning,summarizing, and analyzing financial data. The workflow includes:
##### - Importing and validating CSV data
##### - Removing duplicate records
##### - Generating grouped summary statistics  by business state
##### - Detecting negative financial ratios
##### - Creating a debt-to-income metric
##### - Visualizing financial trends across states

##### Algorithm documentation is provided through both a flowchart and pseudocode
##### to clearly communicate the data processing logic.

## Dataset Description

##### The dataset contains financial metrics for businesses across multiple states,including revenue, equity, and debt variables. These metrics allow us to evaluate financial performance and calculate derived ratios such as debt-to-income.

##### Key variables include:
##### - `Business.State`
##### - `Total.Revenue`
##### - `Total.Equity`
##### - `Total.Long.term.Debt`
##### - `Debt.to.Equity`

##### Flowchart

```{r flowchart, echo=FALSE}
grViz("
digraph task1_flowchart {

  graph [
    layout = dot,
    rankdir = TB,
    bgcolor = white,
    nodesep = 0.5,
    ranksep = 0.8
  ]

  node [
    shape = box,
    style = filled,
    fillcolor = lightblue,
    color = black,
    fontname = Helvetica,
    fontsize = 14,
    margin = 0.15
  ]

  edge [
    color = black,
    penwidth = 1.2,
    fontname = Helvetica,
    fontsize = 12
  ]

  start      [label = 'Start', shape = oval, fillcolor = lightgreen]
  readcsv    [label = 'Read CSV and set as df']
  dupcheck   [label = 'Check for duplicates', shape = diamond, fillcolor = lightyellow]
  removedups [label = 'Remove duplicates from df']
  groupstate [label = 'Group by state, calculate mean, median, min, and max for numeric columns']
  negcheck   [label = 'Are there negative debt to equity entries?', shape = diamond, fillcolor = lightyellow]
  saveneg    [label = 'Save rows as neg_df']
  keepdf     [label = 'Keep debt to equity entries > 0 in df']
  revcheck   [label = 'Is total revenue = 0?', shape = diamond, fillcolor = lightyellow]
  setna      [label = 'Set debt to income to NA']
  computedti [label = 'Compute debt to income as total long term debt / total revenue and set as d_2_I']
  concatdf   [label = 'Concatenate d_2_I to df, set as final_df']
  printdf    [label = 'Print final_df']
  end        [label = 'End', shape = oval, fillcolor = mistyrose]

  start      -> readcsv
  readcsv    -> dupcheck
  dupcheck   -> removedups [label = 'Yes']
  dupcheck   -> groupstate [label = 'No']
  removedups -> groupstate
  groupstate -> negcheck
  negcheck   -> saveneg [label = 'Yes']
  negcheck   -> keepdf [label = 'No']
  saveneg    -> keepdf
  keepdf     -> revcheck
  revcheck   -> setna [label = 'Yes']
  revcheck   -> computedti [label = 'No']
  setna      -> concatdf
  computedti -> concatdf
  concatdf   -> printdf
  printdf    -> end
}
")
```
## Algorithm Pseudocode

1. Begin program
2. Import CSV data into dataframe `df`
3. Identify duplicate rows
4. Remove duplicate rows
5. Group data by business state
6. Calculate mean, median, min, and max for numeric columns
7. Identify rows where Debt-to-Equity < 0
8. Save those rows as `neg_df`
9. Create debt-to-income variable
10. Handle division-by-zero cases
11. Append new variable to dataset
12. Output final dataframe
13. End program

## Flowchart and Pseudocode Logic

The following table summarizes the algorithm used in the data processing workflow.  
Each step corresponds to a component of the flowchart and explains the logical
operations performed during the analysis.

| Flowchart Step | Pseudocode | Explanation |
|----------------|------------|-------------|
| **Begin** | Start program | Marks the beginning of the program execution. |
| **Read CSV and set as `df`** | Import comma-separated value data into a dataframe named `df`. | Establishes a structured container for data manipulation and analysis. |
| **Check for duplicate rows** | Identify duplicate rows in `df`. | Detects duplicate records that may distort statistical summaries or calculations. |
| **Duplicates detected?** | If duplicates exist → remove duplicate rows from `df`. | Ensures data integrity by preventing multiple instances of the same record. |
| **No duplicates detected** | Proceed to the next step. | If no duplicates exist, the dataset remains unchanged. |
| **Group data by state** | Group the dataframe by `Business.State`. | Organizes the data into groups so that summary statistics can be calculated by state. |
| **Calculate summary statistics** | For each numeric variable calculate mean, median, minimum, and maximum. | Produces descriptive statistics that summarize the financial characteristics of businesses within each state. |
| **Check for negative debt-to-equity values** | Filter rows where `Debt.to.Equity < 0`. | Identifies potentially problematic financial entries where the ratio is negative. |
| **Negative values detected?** | Save filtered rows to a new dataframe named `neg_df`. | Preserves anomalous values for inspection while allowing the main dataset to continue through the workflow. |
| **Continue processing** | Proceed with additional calculations and transformations. | The cleaned dataset is now prepared for feature engineering and visualization. |

##### The program identifies companies with negative debt-to-equity ratios and stores them in a separate data frame (neg_df). This separation allows for outliers to be analyzed independently 

##### Is total revenue = 0 Yes: Set debt to income to Na. No: Compute debt to income as total long term debt / total revenue	For each row in df, if total revenue is equal to 0 set debt to income to na, otherwise set

##### debt-to-income column to long-term-debt divided by revenue	The debt-to-income ratio is then introduced to a decision structure. Any company with a revenue of 0 will be labeled Na for its debt-to-income ratio, because dividing by 0 is not possible. Otherwise, the ratio is calculated by long-term debt divided by total revenue

##### Concatenate d_2_I to df, set as final_df	Concatenate d_2_I data frame to df	The debt-to-income ratio is concatenated back into the original dataset for easy queries.
##### Print final_df	Output data from final_df  data frame	Print a readable dataframe with all pertinent calculations
#####	End Program

##### The program's logic follows a structured data processing pipeline designed to ensure accuracy, integrity, and analytical validity. 

##### 1.	Import CSV data into a data frame (df). This establishes a structured container for manipulation.
##### 2.	Detect and remove duplicates in the df to ensure data integrity before calculations. This will prevent measures of central tendency and other calculations from being inaccurate later in the program. 
##### 3.	The program then groups the data by state to compute descriptive statistics: mean, median, minimum, and maximum for all numeric variables. This aggregation step is performed before filtering, ensuring that summary statistics reflect the complete dataset rather than a subset. 
##### 4.	Next, the program identifies companies with negative debt-to-equity ratios and stores them in a separate data frame (neg_df). This separation allows for outliers to be analyzed independently without modifying the cleaned primary dataset.
##### 5.	The debt-to-income ratio is then introduced to a decision structure. Any company with a revenue of 0 will be labeled Na for its debt-to-income ratio, because dividing by 0 is not possible. Otherwise, the ratio is calculated by long-term debt divided by total revenue. 
##### 6.	This conditional logic ensures mathematical accuracy and prevents errors. 
##### 7.	Lastly, the debt-to-income ratio is concatenated back into the original dataset for easy queries. 

##### Overall, Flowcharts offer a visual representation
##### that outlines the program's sequence of actions
##### and decision points, facilitating a comprehensive 
##### understanding of its logic and structure. 
##### Meanwhile, pseudocode serves as a bridge 
##### between human-readable language and actual code,
##### articulating the algorithm's logic in a simplified,
##### syntax-free manner. (Western Governors University, n.d.).
##### Using both flowcharts and pseudocode provides 
##### comprehensive algorithmic documentation, 
##### offering multiple perspectives 

```{r}
df <- read.csv("/Users/nicholasjojola/Desktop/MSDADS/analytical programming/D598 Data Set.csv")
```
##### removes duplicate entries
```{r}
df <- df %>% distinct()
```

##### Grouped data and summarized
```{r}
g_data <- df %>%
  group_by(Business.State) %>%
  summarize(across(where(is.numeric), list(mean = ~ mean (.x, na.rm = TRUE),
    median = ~median (.x, na.rm = TRUE),
    min = ~min(.x, na.rm = TRUE),
    max = ~max(.x, na.rm= TRUE))))
print(g_data)
```
##### filtered debt to equity by negative
```{r}
neg_df <- df %>%
  filter(Debt.to.Equity < 0)
print(neg_df)
```
##### Debt to Income creation
```{r}
d_2_I <- df %>%
  mutate(
        debt.to.income = ifelse(
          Total.Revenue == 0, 
          NA_real_, 
          Total.Long.term.Debt/Total.Revenue
        )
      ) %>%
      select(debt.to.income)
```
##### Concatination of debt to income into the original df

```{r}
final_df <- bind_cols(d_2_I, df)
print(final_df)
```
##### Visualization

```{r}
ggplot(final_df, aes(x = Business.State, y = Total.Revenue)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Total Revenue by State",
    x = "State",
    y = "Revenue"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
ggplot(data = final_df, aes(x=Business.State, y=debt.to.income, fill = Business.State)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 2))
  
ggplot(data = final_df, aes(x = Business.State, y = debt.to.income, color = Business.State)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    panel.background = element_rect(fill = "white"))
    
ggplot(data = final_df, aes(x=Total.Long.term.Debt, y=Total.Equity)) +
  geom_point() +
  theme(legend.position = "none") +
  geom_smooth(method = "lm", se = TRUE, color = "red")

ggplot(final_df, aes(x = Total.Long.term.Debt, y = Total.Equity)) +
  geom_point(color = "darkred") +
  theme_minimal() +
  labs(
    title = "Debt vs Equity Relationship",
    x = "Long Term Debt",
    y = "Total Equity"
  )
```

## Conclusion

This project demonstrates a complete analytics pipeline in R,
including data cleaning, transformation, feature engineering,
and visualization.

The use of flowcharts and pseudocode provides clear documentation
of the algorithmic workflow, ensuring that the analysis is both
reproducible and understandable.

This approach reflects best practices in data analytics,
where transparency and documentation are critical components
of reliable data processing.